---
title: "Statistical power"
author: "Federica Grosso 904701"
date: "2024-08-06"
output:
  html_document: default
  pdf_document: default
   
---

# Power analysis
```{r}
library(ggplot2)
library(tidyverse)
library(ggrepel)
```

### Functions

```{r}
explained_variance <- function(data, N)
{
eaf = data$eaf.exposure
MAF <- ifelse(eaf <= 0.5, eaf, 1-eaf)
beta = data$beta.exposure
se =data$se.exposure
R2 = 2 * beta^2 * MAF * (1 - MAF) / (2 * beta^2 * MAF * (1 - MAF) + se^2 * 2 * N * MAF * (1 - MAF))
R_2=sum(R2)
return(R_2)
}

calculate_power_B <- function(b1, n, ratio, sig, rsq) {
  power <- pnorm(sqrt(n*rsq*(ratio/(1+ratio))*(1/(1+ratio)))*b1-qnorm(1-sig/2))*100
  return(power)
}

calculate_power_C <- function(b1, n, sig, rsq){
  power<-pnorm(sqrt(rsq*b1^2*n)-qnorm(1-sig/2))*100
  return(power)
}
```

# Power of MR analysis

## Load of IV

```{r}
path <- "...\\Harm_data_E_GCST90027663_O_MACDEG.txt"
path1 <- "...\\Harm_data_E_GCST90027663_O_AMD2.txt"
path2 <- "...\\Harm_data_E_GCST90027663_O_EarlyAMD.txt"
path3 <- "...\\Harm_data_E_GCST90027663_O_AMD.txt"
data <- read.table(path, sep = "\t")
data <- data[data$mr_keep != FALSE, ]
data1 <- read.table(path1, sep = "\t")
data1 <- data1[data1$mr_keep != FALSE, ]
data2 <- read.table(path2, sep = "\t")
data2 <- data2[data2$mr_keep != FALSE, ]
data3 <- read.table(path3, sep = "\t")
data3 <- data3[data3$mr_keep != FALSE, ]
```

## Calculation of explained variance R^2

```{r}
N=7738
R_2 <- explained_variance(data, N)
R_21 <- explained_variance(data1, N)
R_22 <- explained_variance(data2, N)
R_23 <- explained_variance(data3, N)
# 0.01859125 con GCST90027737
# 0.01858982 con GCST90027663
R_2
R_21
R_22
R_23
```

## Calculation of power

```{r}
n=456348
n1=56637
n2=105248
n3=430221

pwr <- calculate_power_B(b1=0.2, n=456348, ratio=1295/ 455053, sig= 0.05, rsq= R_2);pwr

pwr <- calculate_power_B(b1=0.2, n1, ratio=3685/52952, sig= 0.05, rsq= R_2);pwr

pwr <- calculate_power_B(b1=0.2, n2, ratio=14034/91214, sig= 0.05, rsq= R_2);pwr

pwr <- calculate_power_B(b1=0.2, n3, ratio=11023/419198, sig= 0.05, rsq= R_2);pwr
```

### Binary outcome

```{r}
cat("Power of analysis with", n, "participants: ", pwr, "%")
```

## Plot of effect size vs power

```{r}
# Effect size grid
beta_values <- seq(0.05, 0.60, by=0.05)
```

```{r}
beta_values <- seq(0.05, 0.60, by=0.05)
ratio_values <- c(1295/ 455053, 3685/52952, 14034/91214, 11023/419198)
n_values <- c(456348,56637,105248,430221)
# Inizializza un data frame vuoto per i risultati
results <- data.frame()

# Itera su ciascun valore di beta
for (beta in beta_values) {
  temp_results <- data.frame(
    Beta = beta,
    SampleSize = n_values,
    Ratio = ratio_values,
    Sigma = 0.05,
    Rsq = R_2,
    Power = mapply(calculate_power_B, beta, n_values, ratio_values, 0.05, R_2)
  )
  results <- rbind(results, temp_results)
}
```

```{r}
1295/ 455053
3685/52952
14034/91214
11023/419198
```


### Plot of power of AMD

```{r}
# Filter the data to find y values where x = 0.2
annotate_data <- results %>% filter(Beta == 0.2)

ggplot(results, aes(x = Beta, y = Power, color = as.factor(Ratio))) + 
  geom_point(size = 2.5) + 
  geom_line(lwd = 1) +
  scale_color_manual(name = "Ratio of Cases to Controls", 
                     values = c("#1f77b4", "#ff7f0e", "#2ca02c", "#d62728"), 
                     labels = c("ratio = 0.0028 (GCST90043776 -\nmain analysis)", 
                                "ratio =  0.026 (Finngen study)",
                                "ratio = 0.06959 (GCST90086112)", 
                                "ratio = 0.1539 (GCST010723)")) +
  labs(x = expression(Causal~Effect~(beta)),
       y = "Power (%)") +
  theme_minimal() +
  theme(
      panel.border = element_rect(color = "darkgrey", fill = NA, size = 1),
      axis.line = element_line(color = "darkgrey", size = 0.5),
      axis.ticks = element_line(color = "darkgrey"),
      axis.text.x = element_text(lineheight = 12),
      axis.text.y = element_text(lineheight = 12)
   ) 
  # geom_text(data = annotate_data, 
  #           aes(label = paste0(round(Power, 1), "%")), 
  #           size = 3.5, color = "black")
```

# LACT vs In15

## Load of IV

```{r}
path <- "...\\Harm_data_E_GCST90027488_O_In15.txt"
path1 <-"...\\Harm_data_E_GCST90027488_O_In15.txt"
path2 <- "...\\Harm_data_E_GCST90027488_O_In15_3.txt"
data <- read.table(path, sep = "\t")
data <- data[data$mr_keep != FALSE, ]
data1 <- read.table(path1, sep = "\t")
data1 <- data1[data1$mr_keep != FALSE, ]
data2 <- read.table(path2, sep = "\t")
data2 <- data2[data2$mr_keep != FALSE, ]
```

## Calculation of explained variance R^2

```{r}
N=7738
R_2 <- explained_variance(data,N);R_2
R_21 <- explained_variance(data1,N);R_21
R_22<- explained_variance(data2,N);R_22
```

## Calculation of power

```{r}
n = 11792 # Sample size
n1 = 5368 # Sample size
n2 = 3301 # Sample size
pwr <- calculate_power_C(0.2, n,0.05, R_2)
pwr1 <- calculate_power_C(0.2, n1,0.05, R_21)
pwr2 <- calculate_power_C(0.2, n2,0.05, R_22)
```


### Continuous outcome

```{r}
cat("Power of analysis with ", n, "participants: ",
    pwr,"%")
cat("\nPower of analysis with ", n1, "participants: ",
    pwr1,"%")
cat("\nPower of analysis with ", n2, "participants: ",
    pwr2,"%")
```

## Plot of power of In15

```{r}
beta_values <- seq(0.05, 0.60, by = 0.05)
n_values <- c(n, n1, n2)
R2_values <- c(R_2, R_21, R_22)

# Inizializza un data frame vuoto per i risultati
results <- data.frame()

# Itera su ciascun valore di beta
for (beta in beta_values) {
  # Itera su ciascuna combinazione di n_values e R2_values
  for (i in 1:length(n_values)) {
    # Calcola la potenza per la combinazione corrente di Beta, SampleSize e Rsq
    power_value <- calculate_power_C(beta, n_values[i], 0.05, R2_values[i])
    
    # Crea un data frame temporaneo per salvare i risultati
    temp_results <- data.frame(
      Beta = beta,
      SampleSize = n_values[i],
      Sigma = 0.05,
      Rsq = R2_values[i],
      Power = power_value
    )
    
    # Aggiungi i risultati temporanei al data frame principale
    results <- rbind(results, temp_results)
  }
}

# Stampa i risultati
print(results)
```

```{r}
# Filter the data to find y values where x = 0.2
annotate_data <- results %>% filter(Beta == 0.2)

ggplot(results, aes(x = Beta, y = Power, color = as.factor(SampleSize))) + 
  geom_point(size = 2.5) + 
  geom_line(lwd = 1) +
  scale_color_manual(name = "Outcome's GWAS accession", 
                     values = c("#1f77b4", "#ff7f0e", "#2ca02c"), 
                     labels = c("GCST90241594 (n=3,301)", 
                                "GCST90087746 (n=5,368)",
                                "GCST90274800 (n=11,792) - \nmain analysis")) +
  labs(x = expression(Causal~Effect~(beta)),
       y = "Power (%)") +
  theme_minimal() +
  theme(
      panel.border = element_rect(color = "darkgrey", fill = NA, size = 1),
      axis.line = element_line(color = "darkgrey", size = 0.5),
      axis.ticks = element_line(color = "darkgrey"),
      axis.text.x = element_text(lineheight = 12),
      axis.text.y = element_text(lineheight = 12)
  ) 
```

# LACT vs TRAIL 

## Load of IV

```{r}
path <- "...\\Harm_data_E_GCST90027488_O_GCST90274843.txt"
path1 <- "...\\Harm_data_E_GCST90027488_O_TRAIL.txt"
data <- read.table(path, sep = "\t")
data <- data[data$mr_keep != FALSE, ]
data1 <- read.table(path1, sep = "\t")
data1 <- data1[data1$mr_keep != FALSE, ]
```

## Calculation of explained variance R^2

```{r}
N=7738
R_2 <- explained_variance(data,N); R_2
R_21 <- explained_variance(data1,N); R_21
```

## Calculation of power

```{r}
n = 14735 # Sample size
n1 = 21758 # Sample size
pwr <- calculate_power_C(0.2, n, 0.05, R_2)
pwr1 <- calculate_power_C(0.2, n1, 0.05, R_21)

cat("Power of analysis with ", n, "participants: ",
   pwr,"%")
cat("\nPower of analysis with ", n1, "participants: ",
     pwr1,"%")
```

## Plot of power of TRAIL

```{r}
beta_values <- seq(0.05, 0.60, by = 0.05)
n_values <- c(n, n1)
R2_values <- c(R_2, R_21)

# Inizializza un data frame vuoto per i risultati
results <- data.frame()

# Itera su ciascun valore di beta
for (beta in beta_values) {
  # Itera su ciascuna combinazione di n_values e R2_values
  for (i in 1:length(n_values)) {
    # Calcola la potenza per la combinazione corrente di Beta, SampleSize e Rsq
    power_value <- calculate_power_C(beta, n_values[i], 0.05, R2_values[i])
    
    # Crea un data frame temporaneo per salvare i risultati
    temp_results <- data.frame(
      Beta = beta,
      SampleSize = n_values[i],
      Sigma = 0.05,
      Rsq = R2_values[i],
      Power = power_value
    )
    
    # Aggiungi i risultati temporanei al data frame principale
    results <- rbind(results, temp_results)
  }
}

# Stampa i risultati
print(results)
```

```{r}
# Filter the data to find y values where x = 0.2
annotate_data <- results %>% filter(Beta == 0.2)

ggplot(results, aes(x = Beta, y = Power, color = as.factor(SampleSize))) + 
  geom_point(size = 2.5) + 
  geom_line(lwd = 1) +
  scale_color_manual(name = "Outcome's GWAS accession", 
                     values = c("#1f77b4", "#ff7f0e"), 
                     labels = c("GCST90274843 (n=11,792) - \nmain analysis", 
                                "GCST90012011 (n=21,758)")) +
  labs(x = expression(Causal~Effect~(beta)),
       y = "Power (%)") +
  theme_minimal() +
  theme(
      panel.border = element_rect(color = "darkgrey", fill = NA, size = 1),
      axis.line = element_line(color = "darkgrey", size = 0.5),
      axis.ticks = element_line(color = "darkgrey"),
      axis.text.x = element_text(lineheight = 12),
      axis.text.y = element_text(lineheight = 12)
  ) 
```

# Bifidobacterium adolescentis vs TNFSF12 

## Load IV

```{r}
path <-"...\\Harm_data_E_GCST90027754_O_GCST90274846.txt" # 1 vs 1
path1 <- "...\\Results_ado_TNFSF12\\Harm_data_E_GCST90032220_O_TNFSF12.txt" # 2 vs 1
path2 <- "...\\Results_GCST90027754_TNFSF12\\Harm_data_E_GCST90027754_O_TNFSF12.txt" # 1 vs 2
data <- read.table(path, sep = "\t")
data <- data[data$mr_keep != FALSE, ]
data1 <- read.table(path1, sep = "\t")
data1 <- data1[data1$mr_keep != FALSE, ]
data2 <- read.table(path2, sep = "\t")
data2 <- data2[data2$mr_keep != FALSE, ]
```

## Calculation of explained variance R^2

```{r}
N=7738
R_2 <-explained_variance(data,N); R_2
N1=5959
R_21 <-explained_variance(data1,N1); R_21
R_22<-explained_variance(data2,N); R_22
```

## Calculation of power

```{r}
n = 14736 # Sample size
n1 =14736 # Sample size
n2 = 3301 # Sample size
pwr <- calculate_power_C(0.2,n,0.05,R_2)
pwr1 <- calculate_power_C(0.2,n1,0.05,R_21)
pwr2 <- calculate_power_C(0.2,n2,0.05,R_22)
```


### Continuous outcome

```{r}
cat("Power of analysis with ", n, "participants: ",
    pwr,"%")
cat("\nPower of analysis with ", n1, "participants: ",
      pwr1,"%")
cat("\nPower of analysis with ", n2, "participants: ",
      pwr2,"%")
```

## Plot of power of TNFSF12

```{r}
beta_values <- seq(0.05, 0.60, by = 0.05)
n_values <- c(n, n1, n2)
R2_values <- c(R_2, R_21, R_22)
ex_o <-c("e1-o1","e2-o1","e1-o2")
a <-1:3

# Inizializza un data frame vuoto per i risultati
results <- data.frame()

# Itera su ciascun valore di beta
for (beta in beta_values) {
  # Itera su ciascuna combinazione di n_values e R2_values
  for (i in 1:length(n_values)) {
    # Calcola la potenza per la combinazione corrente di Beta, SampleSize e Rsq
    power_value <- calculate_power_C(beta, n_values[i], 0.05, R2_values[i])
    
    # Crea un data frame temporaneo per salvare i risultati
    temp_results <- data.frame(
      Beta = beta,
      SampleSize = n_values[i],
      Sigma = 0.05,
      Rsq = R2_values[i],
      Power = power_value,
      Ex_O = ex_o[i]
    )
    
    # Aggiungi i risultati temporanei al data frame principale
    results <- rbind(results, temp_results)
  }
}

# Stampa i risultati
print(results)
```

```{r}
# Filter the data to find y values where x = 0.2
annotate_data <- results %>% filter(Beta == 0.2)

ggplot(results, aes(x = Beta, y = Power, color = as.factor( Ex_O ))) + 
  geom_point(size = 2.5) + 
  geom_line(lwd = 1) +
  scale_color_manual(name = "Exposure-Outcome pairs", 
                     values = c("#1f77b4", "#ff7f0e", "#2ca02c"), 
                     labels = c("Exposure: GCST90027754\nOutcome: GCST90243141\n(main analysis)", "Exposure: GCST90027754\nOutcome: GCST90274846", "Exposure: GCST90032220\nOutcome: GCST90243141")) +
  labs(x = expression(Causal~Effect~(beta)),
       y = "Power (%)") +
  theme_minimal() +
  theme(
      panel.border = element_rect(color = "darkgrey", fill = NA, size = 1),
      axis.line = element_line(color = "darkgrey", size = 0.5),
      axis.ticks = element_line(color = "darkgrey"),
      axis.text.x = element_text(lineheight = 12),
      axis.text.y = element_text(lineheight = 12)
  ) + theme(legend.spacing.y = unit(12, 'pt'), legend.key.size = unit(2, 'lines'))
```
